---
title: "PS270 Final Project"
author: "Will Kugler"
description: "My final project"
output:
  distill::distill_article:
    self_contained: false
---

```{r}
library(tidyverse)
library(WDI)
```


```{r}
data <- WDI(country = "all", indicator = "NY.GDP.PCAP.KD", start = 2000, end = 2022)
head(data)
```

## Project thoughts

I am interested in exploring data related to how the economy in a certain part of the world influences voting patterns and human rights. It would be interesting to look at a comparison of voting breakdowns and human rights in a poor country in Africa with a successful country in North America.



## Milestone 2: Tentative Research Question/Proposal Paragraph
My research question is "How does economic development influence voting patterns and human rights outcomes across countries from 2000 to 2022?" Do wealthier countries show more democratic voting behavior and stronger human rights protections than poorer countries? My hypothesis is that higher levels of economic development will be associated with more democratic voting patters and more human rights protections. A stronger economy can lead to higher education levels, a stronger middle class, institutional structure, etc. My primary independent variable is economic development in GDP per capita (US dollar). My dependent variables are voting patterns and human rights, measured by polls and laws. There will be significant variation in economic development by country, especially if I look at and compare countries on different continents. A positive relationship between GDP per capita and democratic voting behavior/human rights would support my hypothesis (economy is growing, more human rights and democratic voters). No relationship would be shown if a poorer country performed better on democratic measures or human rights protections than wealthy ones, this would show economic development does not systematically influence political or human rights outcomes.




## Milestone 3: Loading, Cleaning, and Visualizing the Data




```{r}
library(tidyverse)
library(WDI)
library(scales)

raw_data <- WDI (country = c("ET", "CA"), indicator <- c(gdp_per_cap = "NY.GDP.PCAP.KD", social_outcome = "SP.DYN.LE00.IN"), start = 2000, end = 2021, extra = FALSE)


clean_comparison <- raw_data |>
  rename(Year = year) |>
  drop_na(gdp_per_cap, social_outcome) 

ggplot(clean_comparison, aes(x = Year, y = social_outcome, color = country)) + geom_line(size = 1.2) + geom_point() + theme_minimal() + labs(title = "Development Trajectory: Canada vs. Ethiopia", subtitle = "Comparing economic influence on life expectancy (Human Rights Proxy", x = "Year", y = "Life Expectancy (Years)", color = "country", caption = "Data Source: World Development Indicators (WDI)") + theme(legend.position = "bottom", plot.title = element_text(face = "bold"))

```





## Milestone 4: One visualization and one analysis that attempts to answer my research question





```{r}
library(dplyr)
library(knitr)
analysis_table <- clean_comparison |>
  group_by(country) |>
  summarize(`Avg GDP` = mean(gdp_per_cap), `Longevity Increase` = max(social_outcome) - min(social_outcome), `Wealth-Health Correlation` = cor(gdp_per_cap, social_outcome))

knitr:: kable(analysis_table, digits = 2, caption = "Table 1: Economic Correlation with Social Outcomes (2000-2021)")
```

Brief Discussion:
This table provides a fascinating contrast between a developed North American economy and an emerging African economy. Both show a strong statistical link between wealth and health. In the poor country (Ethiopia), the economy acts as a "survival engine" - driving human rights and health improvements. In a wealthy country (Canada), the economy acts as a "maintenance engine" - sustaining high standards but seeing diminishing returns for each dollar.
Because the economy in Ethiopia is such a matter of life and death(as seen by the 0.96 correlation), voters there might prioritize economic growth and stability far more than voters in Canada, where the "Longevity Increase" has slowed down and basic survival is already guaranteed.