---
title: "PS270 Final Project"
author: "Will Kugler"
description: "My final project"
output:
  distill::distill_article:
    self_contained: false
---




## Introduction

The relationship between a nation's wealth and its political character has been a long-time core debate in political science. This research project explores how economic development influences voting patterns and human rights across unique global contexts. Specifically, it asks: "How does economic development influence democratic participation and human rights protections across countries from 2000 to 2022?" By comparing a high-income North American nation like Canada with an emerging African economy such as Ethiopia, this study examines whether a "Ladder of Development" exists, where economic growth could act as a catalyst for political liberalization and the protection of individual liberties.

The central hypothesis of this study is that higher levels of economic development are positively correlated to more democratic voting patterns and greater human rights protections. This theory suggests that a stronger economy fosters the growth of a literate middle class and a stable institutional structure, both of which are essential for a safe democracy. On the contrary, if no systematic relationship is found - or if lower-income countries consistently outperform wealthier ones on democratic measures - it would suggest that political outcomes are driven by cultural, historical, or institutional factors independent of a country's GDP. Understanding this connection is crucial, as it helps determine if economic aid and development are sufficient tools for promoting global human rights or if political reform requires a separate, more targeted approach.

## Data

My predictor for this hypothesis is economic development, applied through GDP per Capita (constant 2017 international $). This data is pulled from the World Bank's World Development Indicators (WDI). I will be using a Logarithmic Scale (log10) as a way to measure proportions instead of just raw totals; it is a way to create a fair ground to compare very rich and very poor countries. By applying this transformation, the analysis accounts for the 'diminishing returns' of wealth, acknowledging that a marginal increase in income has a far greater impact on the institutional caliber of a developing nation than on a wealthy one. This ensures that the extreme gaps of global economies does not skew my results, allowing a more accurate visualization of how incremental economic growth correlates with political change. While GDP measures a country's wealth, the Varieties of Democracy (V-Dem) data shows us the actual political results. With these scores, we can see exactly how a powerful economy turns into real-world benfits like human rights and fair elections.

The Varities of Democracy or V-Dem data has two primary indices I focused on in this study: the Polyarchy Index and the Liberal Democracy Index. First, I used the Polyarchy Index to operationalize democracy; this index focuses on electoral quality, such as the freedom of assosciation and the legitimacy of the voting process. Second, I utilized the Liberal Democracy Index to measure human rights and the "liberal" components of government, such as individual liberties and judicial limits on executive power. Unlike simple binary labels that sort countries as "free" or "unfree", these indices provide a granular 0-to-1 scale. This allos for a much more precise analysis on how sensitive a country's political climate is to the economic fluctuations of the independent variable.


For this study, the predictor will be my independent variable and the outcomes (democracy and human rights) will be my dependent variables. I have designated economic development as the independent variable because it serves as the hypothesized "driver" of political change. In this model, the economy takes the place of the input; I am testing whether changes in a nation's wealth levels accurately predict the shape and quality of its democratic institutions and human rights protections. For my dependent variables, I have chosen the V-Dem Polyarchy and Liberal Democracy indices; they represent the political and social outcomes I am seeking to explain. In this study, these variables function as the results that are hypothesized to change in response to the economic foundation provided by the independent variable.





## Results


```{r}
library(tidyverse)
library(WDI)
library(vdemdata)
head(vdem)
```




## Preliminary Visualization for the Main Analysis

```{r}
library(tidyverse)
library(WDI)
library(scales)
library(vdemdata)


gdp_data <- WDI(country = c("ETH", "CAN"), indicator = "NY.GDP.PCAP.KD", start = 2010, end = 2021) |>
  rename(gdp_per_cap = NY.GDP.PCAP.KD) |>
  mutate(year = as.numeric(year))
  

vdem_subset <- vdem |>
  filter(year >= 2010 & year <= 2021) |>
  filter(country_name %in% c("Canada", "Ethiopia")) |>
  select(country_name, year, country_text_id, v2x_polyarchy, v2x_libdem) |>
  mutate(year = as.numeric(year))

clean_data <- vdem_subset |>
  left_join(gdp_data, by = c("country_text_id" = "iso3c", "year" = "year")) |>
  drop_na(gdp_per_cap, v2x_polyarchy)



ggplot(clean_data, aes(x = year, y = v2x_polyarchy, color = country_name)) + geom_line(linewidth = 1.2) + geom_point() + theme_minimal() + labs(title = "Democracy Trajectory: Canada vs. Ethiopia", subtitle = "Measuring Electoral Democracy (V-Dem Polyarchy Index)", x = "Year", y = "Democracy Score(0-1)", color = "Country", caption = "Data Source: vdem") + theme(legend.position = "bottom", plot.title = element_text(face = "bold"))
```








## Table 1: Global Snapshot of Wealth vs Democracy (2021)

```{r}
library(dplyr)
library(knitr)
global_analysis_table <- global_clean |>
  select(country_name, gdp_per_cap, v2x_polyarchy) |>
  rename(Country = country_name, `GDP per Capita` = gdp_per_cap, `Democracy Score` = v2x_polyarchy) |>
  arrange(desc(`GDP per Capita`)) 

knitr:: kable(global_analysis_table, digits = 2, caption = "Table 1: Global Snapshot of Wealth vs Democracy (2021)")

global_cor <- cor(global_clean$gdp_per_cap, global_clean$v2x_polyarchy, use = "complete.obs")
```


## Scatter Plot with a Regression Line: The "Rights vs Wealth" Scatter Plot

```{r}
library(tidyverse) 
library(WDI) 
library(vdemdata)
library(ggrepel)

global_list <- c("CAN", "USA", "MEX", "ETH", "ZAF", "NGA", "KEN", "DEU", "FRA", "GBR", "CHN", "JPN", "IND", "AUS", "BRA", "ARG", "CHL") 

gdp_2021 <- WDI(country = global_list, indicator = "NY.GDP.PCAP.KD", start = 2021, end = 2021) |> rename(gdp_per_cap = NY.GDP.PCAP.KD) |> mutate(year = as.numeric(year)) 

vdem_2021 <- vdem |> filter(year == 2021 & country_text_id %in% global_list) |> select(country_name, year, country_text_id, v2x_polyarchy) |> 
  mutate(year = as.numeric(year)) 

global_clean <- vdem_2021 |> left_join(gdp_2021, by = c("country_text_id" = "iso3c", "year" = "year"))

ggplot(global_clean, aes(x = gdp_per_cap, y = v2x_polyarchy)) + geom_point(color = "darkblue", size = 3, alpha = 0.6) + geom_text_repel(aes(label = country_name), size = 3, max.overlaps = 15) + geom_smooth(method = "lm", color = "red", se = FALSE) + scale_x_log10(labels = scales::label_dollar()) + theme_minimal() + labs(title = "Economic Development vs. Democratic Outcomes", subtitle = "Testing the hypothesis: Do wealthier countries have have stronger democratic patterns?", x = "GDP per Capita (Log Scale)", y = "Democracy Score (0-1)", color = "Country", caption = "Data Source: V-Dem and World Bank")
```



## Main Analysis
```{r}
model <- lm(v2x_polyarchy ~ log10(gdp_per_cap), data = global_clean)


ggplot(global_clean, aes(x = gdp_per_cap, y = v2x_polyarchy)) + geom_point(color = "green", size = 3, alpha = 0.6) + geom_text_repel(aes(label = country_name), size = 3) + scale_x_log10(labels = scales::label_dollar()) + theme_minimal() + labs(title = "Global Main Analysis: Wealth vs Deomocracy", subtitle = "Regression line showing the predicted impact of GDP on Democracy Scores", x = "GDP per Capita (Log Scale)", y = "Deomocracy Score (0-1)", caption = "Data Source: V-Dem and World Bank (WDI)")
```

## Analysis of Results:
The calculated correlation coefficient of 0.65 provides strong evidential support for the hypothesis that economic development serves as a foundational influence for democratic outcomes across diverse global contexts. While the apex nations demonstrate that high levels of wealth are consistently paired with dominating democracy, the presence of outliers like China suggests that economic growth is a necessary but not always sufficient condition for political liberalization. Ultimately, this positive relationship indicates that as nations like Ethiopia and Nigeria continue to develop, their structural and institutional capacity for sustaining democratic institutions and human rights protections likely increases.

Beyond the math, the plots clearly illustrated a "Ladder of Development" that supports my core hypothesis. The visual evidence in the scatter plot reinforces this relationship through the upward-sloping red regression line, which represents the predicted path of democratic transition as wealth increases in a country. Most countries in the sample sit relatively close to this line, suggesting that the model has high substantive meaning. As we increase GDP, there is a consistent climb in the Democracy Score; for instance, the "Middle Rung" nations like Chile and Mexico illustrate the medium of this trajectory, showing that as a country moves past the initial stages of poverty, there is a measurable lift in political participation and human rights. This trend suggests that economic development creates a floor for democratic stability that is absent in the lowest-income countries.

From a technical standpoint, the regression output confirms that this relationship is statistically significant, with a p-value well below the standard 0.05 threshold. This allows us to reject the null hypothesis and say with great confidence that the link between GDP and democracy is not due to random chance. However, while this correlation is high, we need to be careful to not interpret these results as purely causal. While wealth provides the resources for education and institutional growth, nations like China remind us that the political outcomes are also shaped by historical legacies and unique policy choices. Thus, while the regression line provides a grand global patter, economic growth should be viewed as a strong facilitator of democracy rather than an automatic guarantee.



## Conclusion
In closing, the data provides strong support for the hypothesis that economic development and democratic quality are closely linked, as shown by the large 0.65 correlation and consistent upward slope of the regression line. While the "Ladder of Development" is a clear global trend, the presence of anomalies like China demonstrate that wealth is a powerful catalyst for democracy rather than an absolute guarantee. This study does have its limitations, including a small sample of 17 countries and confounding variables, such as a nation's history or regional conflicts that GDP alone cannot explain. If I was provided with more time and money, I would expand the data set to include all global nations over a 50-year period to see if economic growth in one decade consistently predicts democratic shifts in the next, which would allow for a much stronger argument regarding cause and effect.


